# Simple Regression {#chapter-2}

```{r, echo = F}
button <-  "position: relative; 
            top: -25px; 
            left: 85%;   
            color: white;
            font-weight: bold;
            background: #4B9CD3;
            border: 1px #3079ED solid;
            box-shadow: inset 0 1px 0 #80B0FB"
```

The focus of this course is linear regression with multiple predictors (AKA *multiple regression*), but we start by reviewing regression with one predictor (AKA *simple regression*). 

## An example from NELS {#example-2}
 
```{r, echo=FALSE, results='asis'}
codefolder::bookdown(init = "hide", style = button)
```

Figure \@ref(fig:fig1) shows the relationship between Grade 8 Reading Achievement (percent correct on a reading test) and Socioeconomic Status (SES; a composite measure on a scale from 0-35). The data are a subsample of the 1988 National Educational Longitudinal Survey (NELS; see https://nces.ed.gov/surveys/nels88/). 

```{r fig1, fig.cap = 'Reading Achievement and SES (NELS88).', fig.align = 'center'}
# Load and attach the NELS88 data
load("NELS.RData")
attach(NELS)

# Scatter plot
plot(x = ses, y = achmat08, col = "#4B9CD3", ylab = "Reading Achievement (Grade 8)", xlab = "SES")

# Run the regression model
mod <- lm(achmat08 ~ ses)

# Add the regression line to the plot
abline(mod) 
```

The strength and direction of the linear relationship between the two variables is summarized by their correlation (specifically, the Pearson product moment correlation). In this sample, the correlation is

```{r}
options(digits = 2)
cor(achmat08, ses)
```

This is a moderate, positive correlation between Reading Achievement and SES. This correlation means that eighth graders from more well-off families (higher SES) also tended to do better in reading (higher Reading Achievement). 

This relationship has been widely documented and discussed in education research (e.g., https://www.apa.org/pi/ses/resources/publications/education). Please share your thoughts on this relationship in class. 

<!-- In terms of Figure \@ref(fig:fig1), the magnitude (strength) of the correlation can be interpreted in terms of how close the points are to the line. If the correlation was exactly one, all of the points would be on the line. If the correlation was zero, the points would be scattered without any (linear) trend at all. See the review materials in Section \@ref(Review) for more info about correlation. -->
